---
title: "Frequently asked questions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frequently asked questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Why isn't my package being snapshotted into the lockfile?

For a package to be recorded in the lockfile, it must be both:

1.  Installed your project library, *and*

2.  Used by the project, as determined by `renv::dependencies()`.

This ensures that only the packages you truly require for your project will enter the lockfile; development dependencies (e.g. `devtools`) normally should not.

So if you find a package is not entering the lockfile, check the output of `renv::dependencies()`.
If an expected package is not listed, it's likely because `dependencies()` uses static analysis and does not understand all of the different ways in which a package might be used in a project.
See the docs for more details.

### Capturing all dependencies

If you'd instead prefer to capture *all* packages installed into your project library (and eschew dependency discovery altogether), you can do so with:

```{r eval=FALSE}
renv::settings$snapshot.type("all")
```

Packages can also be explicitly ignored through a project setting, e.g. with:

```{r eval=FALSE}
renv::settings$ignored.packages("<package>")
```

You might also want to double-check the set of ignored packages (`renv::settings$ignored.packages()`) and confirm that you aren't unintentionally ignoring a package you actually require.

See the documentation in `?snapshot` for more details.

### Capturing explicit dependencies

If you'd like to explicitly declare which packages your project depends on, you can do so by telling renv to form "explicit" snapshots:

```{r eval=FALSE}
renv::settings$snapshot.type("explicit")
```

In this mode, renv will only include packages which are explicitly listed in the project's `DESCRIPTION` file as dependencies.

## How do I update the lockfile?

The most important thing to remember is that `renv::snapshot()` captures the state of your project at the point in time when `renv::snapshot()` was called.
In that sense, the "right" way to update the lockfile is to:

1.  Load the renv project,
2.  Make the changes you want; e.g. install packages; call `options(repos = <...>)`; ...
3.  Call `renv::snapshot()` to update the lockfile.

That said, you are also free to modify the `renv.lock` lockfile by hand if necessary; e.g. if you want to manually add / change repositories, change the version of a package used, and so on.
The `renv.lock` lockfile is a [JSON](https://www.json.org/json-en.html) file, and while no schema is provided, you should be able to infer the structure from the existing fields.

The main downside to editing a package record in the lockfile directly is that you won't be able to provide a `Hash` for that package, and so renv won't be able to use its global package cache when installing that package.

## How should I handle development dependencies?

This is related to the above question: by design, `renv.lock` normally only captures build-time or deploy-time dependencies; it may not capture the packages that you use in iterative workflows (e.g. `devtools`).
However, you may want some way of still ensuring these development dependencies get installed when trying to restore a project library.

For cases like these, we recommend tracking these packages in a project DESCRIPTION file; typically, within the `Suggests:` field.
Then, you can execute:

```{r eval=FALSE}
renv::install()
```

to request that renv install the packages as described in the DESCRIPTION file.
In addition, the `Remotes:` fields will be parsed and used, to ensure packages are installed from their declared remote source as appropriate.

## I'm returning to an older renv project. What do I do?

Suppose you were using renv to manage an older project's dependencies.
You have an older lockfile, capturing the dependencies in use when you were last working with that project.
You now need to resume work on this project -- what do you do?

The answer depends on how exactly you want to use the project.
Do you want to treat it as a "time capsule", with dependencies frozen in time?
Or are the dependencies in this project fluid, and you are primarily using renv just for isolation of project dependencies?

For time capsules, use `renv::restore()` to reinstall the exact packages as declared in the project lockfile `renv.lock`.
You may also need to find and install the older version of R used previously with that project, unless your intention is to upgrade R.

For projects with fluid dependencies, call `renv::update()` to get the latest versions of the dependencies.
Once you've verified that the code still works (or made the changes needed to get it working), call `renv::snapshot()` to record the latest versions.

You can also take a more managed approach, that's somewhat in between the two extremes:

1.  Use `renv::restore()` to restore the project state as defined in the lockfile.

2.  Install and update packages deliberately with `renv::install()` and friends.

3.  Verify your code works, then call `renv::snapshot()` to update the new lockfile.

## Why are package downloads failing?

Some issues ultimately boil down to a lack of connectivity between your machine and the R package repositories and remote sources you are trying to use.
If you are working in a corporate environment, it may be worth confirming whether you have a corporate proxy in place inhibiting internet access, or whether R and renv need to be configured in a way compatible with your working environment.
This is often true on Windows machines in enterprise environments, where the default "wininet" download method may work more reliably than others.
Learn more in `vignette("package-install")`.

In addition, note that renv places shims on the R search path that re-routes calls from `install.packages()` to `renv::install()`.
If you need to bypass these shims, you can use `utils::install.packages(<...>)`; that is, with the call to `install.packages()` explicitly qualified with the package `utils::`.
See `?renv::load` more details.
